---
title: "ASReml vs bayesreml: Dual-Backend Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ASReml vs bayesreml: Dual-Backend Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
HAS_ASREML <- requireNamespace("asreml", quietly = TRUE)
HAS_BAYESREML <- requireNamespace("bayesreml", quietly = TRUE)
```

## Overview

funcrop supports two estimation backends:

| Feature | ASReml (REML) | bayesreml (Bayesian) |
|---------|:------------:|:-------------------:|
| Estimation | Restricted maximum likelihood | MCMC via greta/TensorFlow |
| Speed | Fast (seconds) | Slow (minutes) |
| Licence | Commercial | Open source (GPL-3) |
| Uncertainty | Approx. SE | Full posterior distributions |
| Boundary VCs | Can be zero | Always positive (prior) |
| Model comparison | AIC/BIC | WAIC/LOO-IC |

## Check Available Engines

```{r engines}
library(funcrop)
cat("Available engines:", paste(funcrop_engines(), collapse = ", "), "\n")
cat("Default engine:", funcrop_default_engine(), "\n")
```

## Example: Baseline RCBD Model

We compare both backends on a simple variety trial:

```{r data}
library(data.table)
data(sim_grain_fill)
dt <- copy(sim_grain_fill)
dt[, variety := factor(variety)]
dt[, block := factor(block)]

# Add plot-level yield noise
set.seed(42)
pids <- unique(dt$plot_id)
dt <- merge(dt,
            data.table(plot_id = pids,
                       yield_noise = rnorm(length(pids), 0, 0.8)),
            by = "plot_id", sort = FALSE)
dt[, yield_plot := yield + yield_noise]
yp <- unique(dt[, .(plot_id, variety, block, yield_plot)])
```

### ASReml

```{r asreml, eval = HAS_ASREML}
library(asreml)
m_asr <- asreml(yield_plot ~ block, random = ~ variety,
                data = yp, trace = FALSE)
cat("Converged:", m_asr$converge, "\n\n")
cat("Variance components:\n")
print(summary(m_asr)$varcomp)
```

```{r asreml-placeholder, eval = !HAS_ASREML, echo = FALSE}
cat("ASReml not available. Install from https://vsni.co.uk/software/asreml-r\n")
```

### bayesreml

```{r bayesreml, eval = HAS_BAYESREML}
library(bayesreml)
m_bay <- bayesreml(yield_plot ~ block, random = ~ variety,
                   data = copy(yp),  # copy() prevents by-ref modification
                   n_samples = 500, warmup = 250, chains = 2,
                   verbose = FALSE, mcmc_verbose = FALSE)
cat("Variance components (posterior):\n")
print(m_bay$extras$variance_comps)
```

```{r bayesreml-placeholder, eval = !HAS_BAYESREML, echo = FALSE}
cat("bayesreml not available. See the bayesreml package documentation.\n")
```

## Comparison

```{r compare, eval = HAS_ASREML && HAS_BAYESREML}
# ASReml: variances
asr_sigv <- summary(m_asr)$varcomp["variety", "component"]
asr_sige <- summary(m_asr)$varcomp["units!R", "component"]

# bayesreml: SDs (square for variance comparison)
bay_vc <- m_bay$extras$variance_comps
bay_sigv_sd <- bay_vc$estimate[grep("variety", bay_vc$component)][1]
bay_sige_sd <- bay_vc$estimate[grep("sigma_e", bay_vc$component)][1]

comp <- data.frame(
  Parameter = c("sigma_variety^2", "sigma_residual^2"),
  ASReml = round(c(asr_sigv, asr_sige), 4),
  bayesreml_var = round(c(bay_sigv_sd^2, bay_sige_sd^2), 4),
  pct_diff = round(100 * (c(bay_sigv_sd^2, bay_sige_sd^2) -
                           c(asr_sigv, asr_sige)) /
                     c(asr_sigv, asr_sige), 1)
)
cat("\nVariance component comparison:\n")
print(comp, row.names = FALSE)
```

## Expected Differences

1. **Variance components**: bayesreml reports SDs with lognormal priors;
   REML gives point estimates that can be zero (boundary). After squaring
   bayesreml SDs, expect 5--15% differences for small datasets.

2. **BLUPs vs posterior means**: Rankings are typically identical
   (r > 0.99). Bayesian estimates are slightly more shrunk due to prior.

3. **Convergence**: ASReml may fail for complex models (AI singularity);
   bayesreml with priors often succeeds where REML fails.

4. **Computational cost**: ASReml is 10--100x faster. Use ASReml for
   exploration, bayesreml for final inference with full uncertainty.

5. **bayesreml modifies data.table by reference**: Always use `copy()`
   when passing data.table objects to bayesreml.

## Recommendation

- **Routine analysis**: Use ASReml (fast, well-tested)
- **Publication-quality inference**: Run bayesreml for credible intervals
- **Model diagnostics**: Compare both -- divergence signals weak
  identification or prior sensitivity
- **No ASReml licence**: bayesreml provides full functionality (slower)
