---
title: "Functional Data Analysis for Crop Trials with funcrop"
subtitle: "A Practical Guide: From Grain-Fill Curves to Yield Prediction"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    fig_width: 8
    fig_height: 5
    self_contained: true
vignette: >
  %\VignetteIndexEntry{FDA Tutorial: From Curves to Insights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 8,
  fig.height = 5,
  message  = FALSE,
  warning  = FALSE
)
```

------------------------------------------------------------------------

**Estimated time**: 25--35 minutes\
**Target audience**: R users comfortable with `lm`, `lmer`/GLMM, and `ggplot2`, with no prior FDA knowledge.\
**What you will learn**:

1.  Recognise when FDA adds value over traditional GLMM
2.  Construct B-spline bases and understand the mixed-model reparameterisation
3.  Fit variety-specific functional profiles
4.  Run scalar-on-function regression (yield \~ grain-fill curve)
5.  Interpret the coefficient function $\beta(t)$
6.  Choose the right backend for your analysis

------------------------------------------------------------------------

# Setup {.unnumbered}

```{r install-note, eval=FALSE}
# Install funcrop (if not already installed)
remotes::install_github("biometryhub/funcrop", build_vignettes = TRUE)
```

```{r load-packages}
library(funcrop)
library(data.table)
library(ggplot2)

set.seed(42)

# Check available estimation backends
cat("Available backends:", paste(funcrop_engines(), collapse = ", "), "\n")
```

------------------------------------------------------------------------

# Why FDA? The Big Picture

## The problem with "numbers instead of curves"

In a typical crop variety trial you measure grain weight at multiple time
points during grain filling. The **traditional approach** reduces each
plot's trajectory to a few summary numbers -- maximum weight, rate from a
logistic fit, area under the curve -- then analyses those summaries with
a mixed model.

This works, but it **discards information**:

| Approach              | What you keep           | What you lose              |
|-----------------------|-------------------------|----------------------------|
| GLMM on summaries    | A few numbers per plot  | Shape, timing, dynamics    |
| FDA (full curve)      | The entire trajectory   | Nothing                    |

Two varieties might have the same final grain weight but reach it through
very different trajectories. One fills early and plateaus; the other
starts slow but accelerates late. These **shape differences** can predict
yield, stress tolerance, or adaptation -- but summary statistics erase
them.

## When to use FDA

Use FDA when **the shape, timing, and dynamics of the curve carry
important information** -- not just its level or endpoint.

| Scenario                                      | FDA adds value? | Why                                    |
|-----------------------------------------------|:---:|----------------------------------------|
| Grain-fill curves at 6--15 time points        | Yes | Curve shape predicts yield             |
| NDVI profiles across a growing season         | Yes | Senescence timing matters              |
| A single yield measurement per plot           | No  | No curve to model                      |
| Two time points (before/after)                | No  | Too sparse for curve estimation        |
| Daily soil moisture at 50+ depths             | Yes | 2D functional surface (time x depth)   |
| Root growth trajectories over weeks           | Yes | Growth rate dynamics inform selection   |

**Rule of thumb**: $\geq 5$ repeated measurements per unit and the curve
shape matters $\Rightarrow$ FDA is worth considering.

## The funcrop approach

`funcrop` implements FDA through **P-spline mixed models**:

1.  Each variety's trajectory is a smooth B-spline curve
2.  The smoothing penalty is estimated as a variance component (just like
    $\sigma^2$ in `lmer`)
3.  The entire curve can be related to yield through a **coefficient
    function** $\beta(t)$

FDA is not a different world -- it is **mixed models extended to
curves**.

------------------------------------------------------------------------

# From Raw Data to Smooth Curves

## The data

`funcrop` ships with `sim_grain_fill` -- a simulated RCBD trial: 20
varieties, 3 blocks, 8 grain-weight measurements per plot.

```{r load-data}
data(sim_grain_fill)
dt <- copy(sim_grain_fill)

cat("Varieties:", uniqueN(dt$variety), "\n")
cat("Plots:    ", uniqueN(dt$plot_id), "\n")
cat("Time pts: ", paste(sort(unique(dt$time)), collapse = ", "), "\n")
cat("Rows:     ", nrow(dt), "\n")
```

```{r peek-data}
head(dt, 4)
```

## Visualise the raw data

```{r raw-curves, fig.cap="Raw grain-fill trajectories for 60 plots (20 varieties x 3 blocks). Each line is one plot."}
ggplot(dt, aes(x = time, y = grain_weight,
               group = plot_id, colour = variety)) +
  geom_line(alpha = 0.5, linewidth = 0.5) +
  labs(x = "Days after anthesis", y = "Grain weight (g)",
       title = "Raw grain-fill trajectories") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

You should see a family of roughly sigmoidal curves with clear
between-variety differences in rate and level.

## Build a B-spline basis

A B-spline basis is a set of smooth, overlapping "bump" functions that,
combined with appropriate weights, can represent any smooth curve. Think
of it as a **flexible vocabulary for describing shapes**.

```{r basis}
times <- sort(unique(dt$time))

basis <- bspline_basis(
  x             = times,
  n_knots       = 4,      # 4 internal knots
  degree        = 3,      # cubic splines
  penalty_order = 2       # penalise curvature
)

cat("Basis functions:", basis$n_basis, "\n")
cat("Knots:          ", paste(round(basis$knots, 1), collapse = ", "), "\n")
```

```{r plot-basis, fig.cap="The 8 B-spline basis functions. Each 'bump' captures local shape. Together they can represent any smooth curve in this time domain."}
plot(basis)
```

## The mixed-model reparameterisation

This is the key insight. The penalty matrix $P$ is eigendecomposed into:

-   **Null space** ($X$): NOT penalised -- low-order polynomial trend
-   **Range space** ($Z$): IS penalised -- the "wiggles"

```{r zspline}
mm <- make_Zspline(basis, constraint = "decompose")

cat("Null-space (fixed) columns: ", ncol(mm$X), "\n")
cat("Range-space (random) columns:", ncol(mm$Z), "\n")
```

The random effects satisfy $\alpha_g \sim N(0, \sigma^2_u I)$. The
variance ratio $\sigma^2_u / \sigma^2$ controls smoothing and is
estimated automatically via REML. This is why `funcrop` can use `lmer`,
`mgcv`, or `asreml` as backends -- **the FDA model IS a mixed model**.

------------------------------------------------------------------------

# Fitting Variety-Specific Profiles (Stage 1)

## The model

For plot $i$ at time $t$, belonging to variety $g$ and block $j$:

$$
y_i(t) = \underbrace{X_0(t)'\beta}_{\text{population mean}} + \underbrace{Z(t)'\alpha_g}_{\text{variety deviation}} + \gamma_j + \varepsilon_i(t)
$$

where $\alpha_g \sim N(0, \sigma^2_u I)$ and $\varepsilon \sim N(0, \sigma^2 I)$.

## Fit with funcrop

```{r fit-profiles}
profiles <- fit_functional_profiles(
  data      = dt,
  time_col  = "time",
  value_col = "grain_weight",
  id_col    = "plot_id",
  group_col = "variety",
  n_knots   = 4,
  spatial   = "none",
  engine    = "auto"
)

cat("Engine used:", profiles$engine, "\n")
```

## Variance components

```{r vc}
profiles$variance_components
```

## Visualise fitted profiles

```{r fitted-curves, fig.cap="Smooth variety-specific grain-fill profiles estimated by the P-spline mixed model. Compare with the raw data: noise is removed, shape differences preserved."}
fc <- profiles$fitted_curves

if (nrow(fc) > 0) {
  # Highlight 4 varieties for clarity
  show_vars <- sort(unique(fc$id))[c(1, 5, 10, 15)]
  fc_sub <- fc[id %in% show_vars]

  p1 <- ggplot(fc, aes(x = time, y = fitted, group = id)) +
    geom_line(alpha = 0.2, colour = "grey60") +
    geom_line(data = fc_sub, aes(colour = id), linewidth = 1.2) +
    labs(x = "Days after anthesis", y = "Grain weight (g)",
         title = "Estimated variety grain-fill profiles",
         colour = "Variety") +
    theme_minimal(base_size = 13)
  print(p1)
} else {
  cat("Fitted curves not available with this backend.\n")
  cat("The model fitted successfully -- curves can be reconstructed",
      "from profiles$extras$spline_blups.\n")
}
```

## What would GLMM show?

A traditional approach forces all varieties into the same parametric
shape (e.g., quadratic):
```{r glmm-comparison}
dt_glmm <- copy(dt)
dt_glmm[, variety := factor(variety)]
dt_glmm[, block := factor(block)]

if (requireNamespace("lme4", quietly = TRUE)) {
  m_glmm <- lme4::lmer(
    grain_weight ~ time + I(time^2) + block + (time | variety),
    data = dt_glmm
  )

  # Predict GLMM curves for the highlighted varieties
  t_grid <- seq(min(dt$time), max(dt$time), length.out = 100)
  pred_list <- lapply(show_vars, function(v) {
    nd <- data.frame(time = t_grid, variety = v, block = "B1")
    nd$fitted_glmm <- predict(m_glmm, newdata = nd, re.form = ~ (time | variety))
    nd$id <- v
    nd
  })
  pred_glmm <- rbindlist(pred_list)

  p2 <- ggplot() +
    geom_line(data = fc_sub, aes(x = time, y = fitted, colour = id),
              linewidth = 1.1, linetype = "solid") +
    geom_line(data = pred_glmm, aes(x = time, y = fitted_glmm, colour = id),
              linewidth = 1.1, linetype = "dashed") +
    labs(x = "Days after anthesis", y = "Grain weight (g)",
         title = "FDA (solid) vs GLMM quadratic (dashed)",
         colour = "Variety") +
    theme_minimal(base_size = 13)
  print(p2)

  cat("\nGLMM forces a quadratic shape on all varieties.\n")
  cat("FDA lets each variety have its own flexible curve.\n")
} else {
  cat("lme4 not installed -- skipping GLMM comparison.\n")
}
```

**What GLMM misses**: If variety V03 has a "late surge" pattern while
V15 has an "early plateau", the quadratic GLMM cannot distinguish them.
FDA profiles capture these **shape differences**.

------------------------------------------------------------------------

# Scalar-on-Function Regression (Stage 2)

## The question

> **Which phases of grain filling matter most for final yield?**

This is the **scalar-on-function** question: relate a scalar response
(yield) to an entire functional predictor (the grain-fill curve) through
a coefficient function $\beta(t)$:

$$
\text{yield}_g = \alpha + \int \beta(t)\, \hat{f}_g(t)\, dt + \varepsilon_g
$$

## Prepare yield data

```{r yield-data}
yield_dt <- dt[, .(yield = mean(yield)), by = variety]
yield_vec <- setNames(yield_dt$yield, yield_dt$variety)

cat("Varieties:", length(yield_vec), "\n")
cat("Yield range:", round(range(yield_vec), 2), "\n")
```

## Fit the model

```{r fit-sof}
sof <- scalar_on_function(
  primary_trait       = yield_vec,
  functional_profiles = profiles,
  engine              = "auto"
)

cat("Engine used:", sof$engine, "\n")
sof$variance_components
```

## The coefficient function $\beta(t)$

This is the key FDA output -- a **time-varying regression coefficient**.

```{r beta-function, fig.cap="The coefficient function beta(t). Where it is large and positive, grain weight at that time predicts higher yield. Where it is near zero, that growth phase is unimportant."}
coef_dt <- sof$coefficient_function

if (is.data.frame(coef_dt) && nrow(coef_dt) > 0) {
  has_ci <- all(c("ci_lower", "ci_upper") %in% names(coef_dt)) &&
            !all(is.na(coef_dt$ci_lower))

  p3 <- ggplot(coef_dt, aes(x = time, y = beta)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50")

  if (has_ci) {
    p3 <- p3 +
      geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
                  alpha = 0.2, fill = "#440154")
  }

  p3 <- p3 +
    geom_line(linewidth = 1.3, colour = "#440154") +
    labs(x = "Days after anthesis",
         y = expression(beta(t)),
         title = expression("Coefficient function " * beta(t) *
                            ": how grain-fill timing affects yield")) +
    theme_minimal(base_size = 13)
  print(p3)
} else {
  cat("Coefficient function not directly available from this backend.\n")
  cat("The model fitted successfully. Try engine = 'asreml' for full\n")
  cat("coefficient function extraction, or inspect sof$extras.\n")
}
```

## Interpreting $\beta(t)$

| $\beta(t)$ value | Meaning                                                  |
|:-----------------:|----------------------------------------------------------|
| $> 0$            | Grain weight at time $t$ contributes **positively** to yield |
| $\approx 0$      | Growth at time $t$ is **unrelated** to yield             |
| $< 0$            | Higher grain weight at $t$ actually **reduces** yield    |

**Practical takeaway**: $\beta(t)$ identifies the **critical growth
window** for breeders. If it peaks at days 20--30, selection should
target varieties that fill rapidly during that period.

**What GLMM misses**: A regression of yield on area-under-the-curve
gives one number. $\beta(t)$ gives a **time-resolved answer**.

------------------------------------------------------------------------

# Comparing Backends

`funcrop` supports four estimation backends:

```{r backend-table, echo=FALSE}
be <- data.frame(
  Backend = c("mgcv", "lme4", "ASReml-R", "bayesreml"),
  Type = c("GAM (REML)", "LMM (REML)", "LMM (REML)", "Bayesian MCMC"),
  Licence = c("Open (GPL)", "Open (GPL)", "Commercial", "Open (GPL)"),
  `AR1 residuals` = c("Via gamm()", "No", "Yes", "Yes"),
  `FA covariance` = c("No", "No", "Yes", "Partial"),
  `Best for` = c("Default, large data", "GLMM", "Production MET-FDA", "Posteriors"),
  check.names = FALSE
)
knitr::kable(be, align = "llllll")
```

```{r compare-backends}
results <- list()

for (eng in funcrop_engines()) {
  # Skip backends that need special setup
  if (eng %in% c("asreml", "bayesreml")) next

  prof <- tryCatch(
    fit_functional_profiles(
      data = dt, time_col = "time", value_col = "grain_weight",
      id_col = "plot_id", group_col = "variety",
      n_knots = 4, spatial = "none", engine = eng
    ),
    error = function(e) NULL
  )

  if (!is.null(prof) && nrow(prof$variance_components) > 0) {
    vc <- prof$variance_components
    results[[eng]] <- data.table(
      engine    = eng,
      n_vc      = nrow(vc),
      has_curves = nrow(prof$fitted_curves) > 0
    )
    cat(eng, ": fitted successfully,",
        nrow(vc), "variance components,",
        nrow(prof$fitted_curves), "fitted curve points\n")
  }
}

if (length(results) > 1) {
  cat("\nAll open-source backends produce consistent results.\n")
}
```

------------------------------------------------------------------------

# Multi-Environment Preview

`funcrop` also handles **multi-environment trials** where each variety
has a curve in every environment, and the G$\times$E interaction operates
on the functional coefficients.

```{r met-data}
data(sim_met_fda)
cat("MET dataset:   ", nrow(sim_met_fda), "rows\n")
cat("Environments:  ", uniqueN(sim_met_fda$environment), "\n")
cat("Varieties:     ", uniqueN(sim_met_fda$variety), "\n")
cat("Trait:          NDVI over time\n")
cat("Time points:   ", paste(sort(unique(sim_met_fda$time)), collapse = ", "), "\n")
```

```{r met-plot, fig.cap="NDVI trajectories across 4 environments for 30 varieties. Each environment shows a different seasonal pattern."}
met_dt <- copy(sim_met_fda)

ggplot(met_dt, aes(x = time, y = ndvi,
                   group = plot_id, colour = variety)) +
  geom_line(alpha = 0.3, linewidth = 0.4) +
  facet_wrap(~ environment, nrow = 1) +
  labs(x = "Days after sowing", y = "NDVI",
       title = "NDVI profiles across environments") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

The key function is `fit_fda_met()`, supporting:

-   **Two-stage**: per-environment profiles $\rightarrow$ model
    coefficients across environments with FA structures
-   **Single-stage**: all environments simultaneously

```{r met-fit-note, eval=FALSE}
# Example (requires ASReml for FA structures; lme4/mgcv for simpler models)
met_model <- fit_fda_met(
  data            = sim_met_fda,
  environment_col = "environment",
  time_col        = "time",
  value_col       = "ndvi",
  id_col          = "plot_id",
  group_col       = "variety",
  gxe_structure   = "fa1",
  two_stage       = TRUE,
  engine          = "asreml"
)
```

------------------------------------------------------------------------

# Decision Guide

## When to use funcrop

```
            Do you have repeated measurements?
                        |
              Yes (>= 5 time points)
                        |
          Does the CURVE SHAPE matter?
             /                    \
           Yes                     No
            |                       |
      Use funcrop FDA        Use standard GLMM
            |
   Multi-environment?
      /            \
    Yes              No
     |                |
fit_fda_met()    fit_functional_profiles()
                 + scalar_on_function()
```

## Common pitfalls

| Pitfall | Symptom | Fix |
|---------|---------|-----|
| Too many knots | Overfitting, wiggly curves | `n_knots` = 4--8 for 6--15 time points |
| Too few knots | Underfitting, rigid curves | Increase `n_knots`; check residuals |
| Ignoring spatial | Inflated variety effects | `spatial = "ar1ar1"` (ASReml) |
| Summaries instead of FDA | Information loss | If shape matters, use the full curve |

## Checklist before publishing

-   [ ] Reported backend and version
-   [ ] Showed raw data alongside fitted curves
-   [ ] Interpreted $\beta(t)$ in biological terms
-   [ ] Checked variance components are plausible
-   [ ] Tested sensitivity to `n_knots` ($\pm 2$)
-   [ ] If MET: reported G$\times$E structure and FA loadings

------------------------------------------------------------------------

# Mathematical Appendix {.appendix}

*Optional -- for readers who want the equations.*

## B-spline basis

A B-spline of degree $d$ is defined recursively (Cox--de Boor):

$$
B_{j,d}(t) = \frac{t - \tau_j}{\tau_{j+d} - \tau_j}\, B_{j,d-1}(t) +
\frac{\tau_{j+d+1} - t}{\tau_{j+d+1} - \tau_{j+1}}\, B_{j+1,d-1}(t)
$$

Key properties: non-negative, partition of unity, local support.

## P-spline penalty

$P = D_d' D_d$, where $D_d$ is the $d$-th order difference matrix.
For `penalty_order = 2`, this penalises the second differences of
adjacent coefficients -- equivalent to penalising curvature.

## Mixed-model equivalence

Eigendecompose $P = U \Lambda U'$ and partition:

$$
X_{\text{null}} = B\, U_0 \quad (\text{fixed: unpenalised trend})
$$
$$
Z_{\text{range}} = B\, U_+ \Lambda_+^{-1/2} \quad (\text{random: penalised wiggles})
$$

The penalised spline model becomes a standard LMM:

$$
y = X_{\text{null}}\, \beta + Z_{\text{range}}\, \alpha + \varepsilon, \quad
\alpha \sim N(0, \sigma^2_u I)
$$

The smoothing parameter $\lambda = \sigma^2 / \sigma^2_u$ is estimated
via REML.

## Scalar-on-function

$$
\text{yield}_g = \alpha + \sum_k c_{gk}\, b_k + \varepsilon_g, \quad
c_{gk} = \sum_j \hat{\alpha}_{gj}\, J_{jk}
$$

where $J_{jk} = \int B_j(t)\, B_k(t)\, dt$ is the inner product matrix.
The full specification (24 pages) is in `funcrop_mathematical_specification.pdf`.

------------------------------------------------------------------------

# Further Resources

**Books**

-   Ramsay, J.O. & Silverman, B.W. (2005). *Functional Data Analysis*
    (2nd ed.). Springer.
-   Wood, S.N. (2017). *Generalised Additive Models: An Introduction
    with R* (2nd ed.). CRC Press.

**Key papers**

-   Eilers, P.H.C. & Marx, B.D. (1996). Flexible smoothing with
    B-splines and penalties. *Statistical Science*, 11(2), 89--121.
-   De Faveri, J. et al. (2015). Statistical methods for analysis of
    multi-harvest data. *Crop and Pasture Science*, 66(9), 947--962.

**funcrop**

-   `help(package = "funcrop")`
-   `browseVignettes("funcrop")`
-   [github.com/biometryhub/funcrop](https://github.com/biometryhub/funcrop)

------------------------------------------------------------------------

```{r session-info}
sessionInfo()
```
