---
title: "Grain-Fill Case Study: FDA for Yield Prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grain-Fill Case Study: FDA for Yield Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Background

In crop variety trials, grain weight is measured over time during the
grain-filling period. The **rate** and **pattern** of grain fill may
influence final yield. Traditional approaches fit parametric curves
(logistic, Gompertz) to each plot and extract parameters (rate, maximum).
FDA provides a more flexible alternative: represent each variety's grain-fill
trajectory as a smooth curve and relate the entire curve to yield.

## Data

We use the simulated `sim_grain_fill` dataset: 20 varieties, 3 blocks,
8 time points (10--45 days after anthesis).

```{r load-data}
library(funcrop)
library(data.table)

data(sim_grain_fill)
dt <- copy(sim_grain_fill)
dt[, variety := factor(variety)]
dt[, block := factor(block)]
dt[, plot_id := factor(plot_id)]

# Add plot-level yield noise (original yield is per-variety)
set.seed(42)
pids <- unique(dt$plot_id)
dt <- merge(dt,
            data.table(plot_id = pids,
                       yield_noise = rnorm(length(pids), 0, 0.8)),
            by = "plot_id", sort = FALSE)
dt[, yield_plot := yield + yield_noise]
```

## Step 1: B-Spline Basis

```{r basis}
times <- sort(unique(dt$time))
basis <- bspline_basis(times, n_knots = 4, degree = 3, penalty_order = 2)
cat("Basis functions:", basis$n_basis, "\n")

# Evaluate at observation times
B_obs <- bspline_basis(dt$time, n_knots = 4, degree = 3,
                        boundary = basis$boundary)$B
```

## Step 2: Per-Plot OLS Curves

Fit each plot's grain-fill trajectory by OLS (no mixed model yet):

```{r ols-curves}
plots <- unique(as.character(dt$plot_id))
alpha_hat <- matrix(NA, nrow = length(plots), ncol = ncol(B_obs))
rownames(alpha_hat) <- plots

for (p in seq_along(plots)) {
  idx <- which(as.character(dt$plot_id) == plots[p])
  y_p <- dt$grain_weight[idx]
  B_p <- B_obs[idx, ]
  alpha_hat[p, ] <- solve(crossprod(B_p), crossprod(B_p, y_p))
}

cat("Fitted", length(plots), "plot-level curves\n")
cat("Coefficient matrix:", nrow(alpha_hat), "x", ncol(alpha_hat), "\n")
```

## Step 3: Variety-Mean Curves

Average the per-plot coefficients within each variety:

```{r variety-means}
pvm <- unique(data.table(plot_id = as.character(dt$plot_id),
                          variety = as.character(dt$variety)))
var_levels <- sort(unique(as.character(dt$variety)))
n_var <- length(var_levels)
n_basis <- ncol(B_obs)

alpha_variety <- matrix(NA, nrow = n_var, ncol = n_basis)
rownames(alpha_variety) <- var_levels

for (v in var_levels) {
  v_plots <- pvm[variety == v, plot_id]
  alpha_variety[v, ] <- colMeans(alpha_hat[v_plots, , drop = FALSE])
}

# Reconstruct curves on a fine grid
t_fine <- seq(min(times), max(times), length.out = 200)
B_fine <- bspline_basis(t_fine, n_knots = 4, degree = 3,
                         boundary = basis$boundary)$B
curves <- B_fine %*% t(alpha_variety)
```

```{r plot-curves, fig.cap = "Variety-mean grain-fill curves (4 selected varieties)."}
show_vars <- var_levels[c(1, 5, 10, 15)]
cols <- c("#440154", "#31688e", "#35b779", "#fde725")

plot(range(t_fine), range(curves[, show_vars]), type = "n",
     xlab = "Days after anthesis", ylab = "Grain weight (g)",
     main = "Variety-Mean Grain-Fill Curves")
for (i in seq_along(show_vars)) {
  lines(t_fine, curves[, show_vars[i]], col = cols[i], lwd = 2)
}
legend("topleft", legend = show_vars, col = cols, lwd = 2, bty = "n")
```

## Step 4: Scalar-on-Function -- yield ~ grain-fill curve

Compute the functional covariate matrix and regress yield on it:

```{r scalar-on-function}
# Inner product matrix J
t_quad <- seq(min(times), max(times), length.out = 500)
B_quad <- bspline_basis(t_quad, n_knots = 4, degree = 3,
                         boundary = basis$boundary)$B
J <- crossprod(B_quad) * diff(t_quad[1:2])

# Functional covariate: C = Alpha %*% J
C_mat <- alpha_variety %*% as.matrix(J)

# Yield data
yield_dt <- dt[, .(yield = mean(yield_plot)), keyby = variety]

# Simple OLS regression: yield ~ C
reg_dt <- data.frame(
  yield = yield_dt$yield[match(var_levels, yield_dt$variety)]
)
for (k in seq_len(n_basis)) {
  reg_dt[[paste0("C", k)]] <- C_mat[, k]
}

fit_ols <- lm(as.formula(paste("yield ~",
               paste0("C", 1:n_basis, collapse = " + "))),
              data = reg_dt)

cat("R-squared:", round(summary(fit_ols)$r.squared, 3), "\n")
cat("RMSE:", round(sqrt(mean(fit_ols$residuals^2)), 3), "\n")
```

## Step 5: The beta(t) Coefficient Function

The regression coefficients `b` define the importance function
beta(t) = sum_k b_k * B_k(t):

```{r beta-t, fig.cap = "Coefficient function beta(t): how each phase of grain-fill influences yield."}
b_hat <- coef(fit_ols)[paste0("C", 1:n_basis)]
beta_t <- as.numeric(B_fine %*% b_hat)

plot(t_fine, beta_t, type = "l", lwd = 2, col = "#440154",
     xlab = "Days after anthesis", ylab = expression(beta(t)),
     main = expression("Coefficient Function " * beta(t)))
abline(h = 0, lty = 2, col = "grey50")
```

**Interpretation**: Where beta(t) is large and positive, grain-fill at that
time contributes positively to yield. Where beta(t) is near zero, that phase
is unimportant. This identifies the critical growth window for breeders.

## Step 6: Penalised beta(t) via Mixed Model

Decompose the regression into fixed (polynomial) and random (wiggly) parts:

```{r penalised}
mm <- make_Zspline(basis, constraint = "decompose")

# Null-space and range-space functional covariates
C_null  <- C_mat %*% mm$X
C_range <- C_mat %*% mm$Z

reg_pen <- data.frame(yield = reg_dt$yield)
for (k in seq_len(ncol(C_null))) {
  reg_pen[[paste0("Cnull", k)]] <- C_null[, k]
}
for (k in seq_len(ncol(C_range))) {
  reg_pen[[paste0("Crange", k)]] <- C_range[, k]
}

cat("Fixed (null-space) covariates:", ncol(C_null), "\n")
cat("Random (range-space) covariates:", ncol(C_range), "\n")
cat("Total observations:", nrow(reg_pen), "\n")
cat("\nNote: With only 20 varieties and", ncol(C_null) + ncol(C_range),
    "covariates,\nthe penalised model requires a mixed model backend",
    "(ASReml or bayesreml)\nto estimate the smoothing parameter. See the",
    "'ASReml vs bayesreml' vignette.\n")
```

## Summary

This vignette demonstrated the funcrop FDA workflow:

1. **B-spline basis** construction with penalty
2. **Per-plot OLS** curve fitting
3. **Variety-mean** functional profiles
4. **Scalar-on-function** regression: yield ~ integral(beta(t) * f(t)) dt
5. **beta(t)** coefficient function revealing which growth phases matter
6. **Penalised estimation** via mixed model reparameterisation

The key advantage over parametric (logistic/Gompertz) approaches: no
assumption about curve shape, and beta(t) reveals the full time-varying
relationship between grain-fill and yield.
