---
title: "Introduction to funcrop"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to funcrop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

**funcrop** provides Functional Data Analysis (FDA) methods for crop variety
trials, integrating B-spline basis functions with linear mixed models. It
supports a dual-backend architecture:

- **ASReml-R** (REML estimation) -- fast, production-proven
- **bayesreml** (Bayesian MCMC via greta) -- full posterior inference

The package handles:

1. **Scalar-on-function regression** -- relating yield to grain-fill curves
2. **Two-dimensional functional traits** -- time x depth, time x wavelength
3. **Multi-environment trials** with GxE interaction on functional traits
4. **Genomic/pedigree integration** for prediction

## Quick Start: B-Spline Basis Construction

The core building block is the B-spline basis:

```{r basis}
library(funcrop)
library(data.table)

# Simulated grain-fill time points
times <- c(10, 15, 20, 25, 30, 35, 40, 45)

# Construct B-spline basis with 4 internal knots, cubic, order-2 penalty
basis <- bspline_basis(times, n_knots = 4, degree = 3, penalty_order = 2)
basis
```

The basis matrix `B` has `r nrow(basis$B)` rows (time points) and
`r basis$n_basis` columns (basis functions). The penalty matrix `P`
enforces smoothness.

```{r plot-basis, fig.cap = "B-spline basis functions evaluated at grain-fill time points."}
plot(basis)
```

## Mixed Model Reparameterisation

The key insight linking P-splines to mixed models: decompose the basis into
a **fixed** (null space) and **random** (range space) component.

```{r zspline}
mm <- make_Zspline(basis, constraint = "decompose")
cat("Fixed (polynomial) columns:", ncol(mm$X), "\n")
cat("Random (wiggly) columns:   ", ncol(mm$Z), "\n")
```

This means a penalised spline can be fitted as a standard mixed model where
the smoothing parameter is estimated as a variance ratio via REML or
Bayesian posterior.

## Simulated Data

funcrop ships with two simulated datasets:

```{r data}
data(sim_grain_fill)
cat("Grain-fill trial:", nrow(sim_grain_fill), "rows\n")
cat("  Varieties:", uniqueN(sim_grain_fill$variety), "\n")
cat("  Blocks:", uniqueN(sim_grain_fill$block), "\n")
cat("  Time points:", sort(unique(sim_grain_fill$time)), "\n")

data(sim_met_fda)
cat("\nMET trial:", nrow(sim_met_fda), "rows\n")
cat("  Varieties:", uniqueN(sim_met_fda$variety), "\n")
cat("  Environments:", uniqueN(sim_met_fda$environment), "\n")
```

## Available Engines

```{r engines}
funcrop_engines()
```

## Genomic Relationship Matrices

funcrop includes utilities for constructing relationship matrices:

```{r grm}
# Small example: 10 varieties, 50 markers
set.seed(123)
markers <- matrix(sample(0:2, 500, replace = TRUE), nrow = 10)
rownames(markers) <- paste0("V", 1:10)

G <- make_genomic_matrix(markers, method = "vanraden1")
cat("GRM dimensions:", nrow(G), "x", ncol(G), "\n")
cat("Mean diagonal:", round(mean(diag(G)), 3), "\n")
```

## Next Steps

See the **Grain-Fill Case Study** vignette for a complete worked example
using `fit_functional_profiles()` and `scalar_on_function()`.
